DOpElib: A Goal Oriented Software Library for Computing PDEs and Optimization Problems

### 2012-11-07 Paper_May_2012/main.tex
Paper: Einleitung etwas ueberarbeitet

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{introduction}
Today, there exist a broad variety of software libraries
that provide numerical tools for solving partial differential
equations (PDEs). Prominent examples are 
\deal{} \cite{dealnew}, dune \cite{dune} , 
or commercial solvers like Ansys \cite{ansys}). 
In fact, these libraries provide very useful tools 
for a huge variety of applications.

However, often, the user is confronted with 
certain bottlenecks. Either, standard high-level routines 
which are necessary to solve even simplest PDEs,
such as linear and nonlinear solvers or timestepping schemes must be implemented 
by the user himself. However, such routines are problem-independent 
and could be provided as standard tools. In fact, in many 
numerical studies, the main task is to compute a  
target functional which represents a certain error norm,
point value, etc.
Second, there is a ever increasing interest in computing 
optimization problems in which the state equation is governed 
by a PDE. Such optimization problems comprise optimal control,
parameter estimation, inverse problems or optimal experimental design. 
So far, there exist a very limited number of open source packages
which provides such routines for scientific studies.

@@

While \deal{} (which might presently seen as the basis because 
it provides various finite elements for discretization) 
leaves much of the work and many
decisions to the user, \dope{} intends to be user-optimized by delivering
prefabricated tools which require from the user only adjustments connected
to his specific problem. The solution of optimization problems with PDE
constraints is an innovation in the \dope{} framework.
The focus is on the numerical solution of both stationary and nonstationary
problems which come from different application fields, like elasticity and
plasticity, fluid dynamics, and fluid-structure interactions.

Summarizing, 
at the present stage the following key features are supported by the library
\begin{itemize}
\item Solution of stationary and nonstationary PDEs in 1d, 2d, and 3d.
\item Various time stepping schemes (based on finite differences), 
  such as forward Euler, backward Euler,
  Crank-Nicolson, shifted Crank-Nicolson, and Fractional-Step-$\Theta$ scheme.
\item All finite elements of from \deal{} including hp-support.
\item Several examples showing the solution of several PDEs including
   Poisson, Navier-Stokes, Plasticity and fluid-structure interaction problems. 
\item Self written line search and trust region newton algorithms for the 
   solution of optimization problems with PDEs \cite{NoWr00}
\item Interface to SNOPT for the solution of optimization problems with PDEs and
  additional other constraints.
\item Several examples showing how to solve various kinds of optimization problems
  involving stationary PDE constraints.
\item Goal-oriented mesh adaptation with the Dual Weighted Residual method, see \cite{BR03}.
\item Different spatial triangulations for control and state variables.
\end{itemize}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{introduction}
Today, there exist a broad variety of software libraries
that provide numerical tools for solving partial differential
equations (PDEs). Prominent examples are 
\deal{} \cite{dealnew}, dune \cite{dune} , 
or commercial solvers like ANSYS \cite{ansys} or COMSOL \cite{comsol}. 
In fact, these libraries provide very useful tools 
for a huge variety of applications.

However, often, the user is confronted with 
certain bottlenecks. Either in open-source projects for academic research, 
standard high-level routines 
which are necessary to solve even simplest PDEs,
such as linear and nonlinear solvers or timestepping schemes must be implemented 
by the user himself. However, such routines are problem-independent 
and could be provided as standard tools. In fact, in many 
numerical studies, the main task is to compute a  
target functional which represents a certain error norm,
point value or traction, for example.
Second, there is nowadays an increasing interest in computing 
optimization problems in which the state equation is governed 
by a PDE. Such optimization problems comprise optimal control,
parameter estimation, inverse problems or optimal experimental design. 
So far, there exist a very limited number of open source packages
which provides standard routines that serve as basis for scientific studies.

@@

While \deal{} (which might presently seen as the basis of DOpE  
various finite elements are taken for the discretization) 
leaves much of the work and many
decisions to the user, \dope{} intends to be user-optimized by delivering
prefabricated tools which require from the user only adjustments connected
to his specific problem. The solution of optimization problems with PDE
constraints is an innovation in the \dope{} framework.
The focus is on the numerical solution of both stationary and nonstationary
problems which come from different application fields, like elasticity and
plasticity, fluid dynamics, and fluid-structure interactions.

The project was initially initiated at the University of Heidelberg in 2010 and  
at the present stage the following key features are supported by the library
\begin{itemize}
\item Solution of stationary and nonstationary PDEs in 1d, 2d, and 3d.
\item Various time stepping schemes (based on finite differences), 
  such as forward Euler, backward Euler,
  Crank-Nicolson, shifted Crank-Nicolson, and Fractional-Step-$\Theta$ scheme.
\item All finite elements of from \deal{} including hp-support.
\item Several examples showing the solution of several PDEs including
   Poisson, Navier-Stokes, Plasticity and fluid-structure interaction problems. 
\item Self written line search and trust region newton algorithms for the 
   solution of optimization problems with PDEs \cite{NoWr00}
\item Interface to SNOPT for the solution of optimization problems with PDEs and
  additional other constraints.
\item Several examples showing how to solve various kinds of optimization problems
  involving stationary PDE constraints.
\item Goal-oriented mesh adaptation with the Dual Weighted Residual method, see \cite{BR03}.
\item Different spatial triangulations for control and state variables.
\end{itemize}
